### 2025-09-23 paper/sec/introduction.tex
Rehash intro a bit.

\figref{scigen-example-website}, generated from our implementation, illustrates these two features. The upper
section shows an excerpt from \cite{zhang18}, a scholarly article comparing text encoding techniques. When a
reader hovers over the phrase \textfrag{does not further improve}, the relevant data are highlighted on the
left. Other fragments (e.g. \textfrag{better than}, \textfrag{further improvements}) that refer to the same
data are also marked, allowing the reader to explore supporting and contrasting evidence. The lower section
shows a counterfactual situation where the authors' experiments had produced different results: here the
phrase \textfrag{does not further improve} is replaced by \textfrag{further improves}.

Transparent documents thus turn static text into interactable, data-driven content able to expose the
evidential basis of scholarly claims, for which we envisage two use cases. First, when \textbf{authoring}
content for an online article, a journalist or scientific publisher may wish to provide text which is linked
to the underlying data so that the evidence base for the textual claims can be explored directly from the
article. Second, when \textbf{reading} a document reporting on findings derived from open data (perhaps a
scientific paper or climate report), the reader may want to retroactively interpret parts of the text as
queries over the available data and gradually ``rationally reconstruct'' the relationship between claims in
the paper and the evidence base. This might be just to aid their own comprehension, or part of a formal peer
review process.

@@

@@

\subsection{Contributions}

\begin{itemize}
    \item Design and proof-of-concept implementation of AI-assisted workflow for authoring transparent text
    (\secref{authoring-workflow})
    \item Empirical evaluation of how effective current LLMs are at providing the ``AI-assisted'' part
\end{itemize}

>>>>

\figref{scigen-example-website}, generated from our implementation, illustrates these two features. The upper
section shows a ``transparent'' excerpt from \cite{zhang18}, a scholarly article comparing text encoding
techniques. When a reader hovers over the phrase \textfrag{does not further improve}, the relevant data are
highlighted on the left. Other fragments (e.g. \textfrag{better than}, \textfrag{further improvements}) that
refer to the same data are also marked, allowing the reader to explore supporting and contrasting evidence.
The lower section shows a counterfactual situation where the authors' experiments had produced different
results: here the phrase \textfrag{does not further improve} is replaced by \textfrag{further improves}.

@@

This transparent version of the document was implemented in Fluid. The source code is shown in
\figref{fluid-example-paragraph}, and makes use of several helper functions, a representative subset of which
are shown in \figref{fluid-scigen}. Both the provenance-tracking runtime of Fluid \emph{and} the LLM-based
authoring support contribute to the solution: the Fluid runtime provides the interactions, and the LLM-based
authoring tool makes the whole process practical. Without tool support, authoring a scientific document as a
(hand-crafted) \emph{program} is unlikely to be a feasible task.

AI-assisted authoring of transparent documents thus support turning static text into interactable, data-driven
content able to expose the evidential basis of scholarly claims, for which we envisage two use cases. First,
when \textbf{authoring} content for an online article, a journalist or scientific publisher may wish to
provide text which is linked to the underlying data so that the evidence base for the textual claims can be
explored directly from the article. Second, when \textbf{reading} a document reporting on findings derived
from open data (perhaps a scientific paper or climate report), the reader may want to retroactively interpret
parts of the text as queries over the available data and gradually ``rationally reconstruct'' the relationship
between claims in the paper and the evidence base. This might be just to aid their own comprehension, or part
of a formal peer review process.

@@
